---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# stocksr

<!-- badges: start -->
[![R-CMD-check](https://github.com/omnia99-hub/stocksr/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/omnia99-hub/stocksr/actions/workflows/R-CMD-check.yaml)
<!-- badges: end -->

The **`stocksr`** R package is designed to facilitate the cleaning, transformation, analysis, and modeling of sector-level stock market data, specifically leveraging S&P 500 daily price data (2010–2020). It is useful for financial data scientists, analysts, and students exploring time-series modeling, technical indicators, and sector-based financial insights.
------------------------------------------------------------------------
## Overview
This package provides a comprehensive framework to:
-   Load and clean raw stock data from Excel files
-   Segment data by market sectors (e.g., Technology, Healthcare)
-   Compute key technical indicators (e.g., MACD, RSI, Bollinger Bands)
-   Construct machine learning models such as Random Forests
-   Compare model performance across different market sectors

### Data Preparation
-   `load_and_preprocess_data()`: Load data and organize by sectors.
-   `calculate_sector_index()`: Aggregate individual stock prices into sector-level index series.

### Feature Engineering
-   `calculate_technical_indicators()`: Compute popular technical indicators (RSI, MACD, etc.) for each sector.
- `prepare_data_for_lstm()`: Format data into 3D sequences for LSTM input.

### Modeling
-   `build_random_forest_model()`: Train a Random Forest model to predict stock movements based on engineered features.


## Installation

You can install the development version of stocksr from [GitHub](https://github.com/) with:

``` r
# install.packages("pak")
pak::pak("omnia99-hub/stocksr")
```

## Example

```{r }
devtools::load_all(".")
library(stocksr)

```

```{r}
library(ggplot2)
library(dplyr)
library(readxl)
```

# Step 1: Load and preprocess raw CSV data
```{r}
data(df_cleaned2)
sectors <- load_and_preprocess_data(df_cleaned2)
head(sectors)
```


# Step 2: Calculate sector index for Technology
```{r}
health_care <- calculate_sector_index(sectors[["Health Care"]])
head(health_care)
```


# Step 3: Generate technical indicators
```{r}
features <- calculate_technical_indicators(health_care)
head(features)
```

# Step 4: Train a Forecast random forest model
```{r}
results <- build_random_forest_model(features, sector_name = "Health Care")
```


