---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# stocksr

<!-- badges: start -->
[![R-CMD-check](https://github.com/omnia99-hub/stocksr/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/omnia99-hub/stocksr/actions/workflows/R-CMD-check.yaml)
<!-- badges: end -->

The **stocksr** R package provides tools for cleaning, transforming, analyzing, and modeling sector-level stock market data, specifically focused on the Energy and Healthcare sectors using S&P 500 daily price data (2010–2020). It supports financial data analysis, time-series forecasting, and feature engineering for sector-specific insights.
------------------------------------------------------------------------
## Overview
This package provides a comprehensive framework to:
-   Load and clean raw stock data from Excel files
-   Segment data by market sectors (e.g., Energy, Healthcare)
-   Compute key technical indicators (e.g., MACD, RSI, Bollinger Bands)
-   Construct machine learning models such as Random Forests
-   Compare model performance across different market sectors

### Data Preparation
-   `load_and_preprocess_data()`: Load data and organize by sectors.
-   `calculate_sector_index()`: Aggregate individual stock prices into sector-level index series.

### Feature Engineering
-   `calculate_technical_indicators()`: Compute popular technical indicators (RSI, MACD, etc.) for each sector.
- `prepare_data_for_lstm()`: Format data into 3D sequences for LSTM input.

### Modeling
-   `build_random_forest_model()`: Train a Random Forest model to predict stock movements based on engineered features.
- `build_lstm_model()`: Train a LSTM to predict stock movements based on engineered features.


## Installation

You can install the development version of stocksr from [GitHub](https://github.com/) with:

``` r
# install.packages("pak")
pak::pak("omnia99-hub/stocksr")
```

## Example

```{r }
devtools::load_all(".")
library(stocksr)

```

```{r}
library(ggplot2)
library(dplyr)

```

# Step 1: Load and preprocess raw CSV data
```{r}
data(df_cleaned2)
sectors <- load_and_preprocess_data(df_cleaned2)
head(sectors$Healthcare,5)
head(sectors$Energy,5)
```


# Step 2: Calculate sector index 
```{r}
health_care <- calculate_sector_index(sectors[["Healthcare"]])
head(health_care)
energy <- calculate_sector_index(sectors[["Energy"]])
head(energy)
```


# Step 3: Generate technical indicators
```{r}
features_healthcare <- calculate_technical_indicators(health_care)
head(features_healthcare)

features_energy <- calculate_technical_indicators(energy)
head(features_energy)

```

# Step 4: Train the random forest model
```{r}
results_healtcare <- build_random_forest_model(features_healthcare, sector_name = "Healthcare")
results_healtcare$prediction_plot
results_healtcare$importance_plot

results_energy <- build_random_forest_model(features_energy, sector_name = "Energy")
results_energy$prediction_plot
results_energy$importance_plot


```
# Step 5: Train the LSTM model
```{r}
lstm_results_healthcare <- build_lstm_model(health_care, sector_name = "Healthcare")
lstm_results_energy <- build_lstm_model(energy, sector_name = "Energy")
```

##Notes

Input data must have daily stock prices with appropriate company names.

The calculate_technical_indicators function generates lagged features, moving averages, volatility, and MACD indicators needed for modeling.

build_lstm_model and build_random_forest_model internally handle training and performance evaluation.

##Output

Each modeling function returns:

Root Mean Squared Error (RMSE)

Mean Absolute Error (MAE)

R-squared (R²)

Trained model object

##Conclusion

The stocksr package provides a fast and easy workflow to simulate sector-level stock forecasting, focusing only on Energy and Healthcare sectors.

## Information 

* The code in this repository was written by Shalu Shalu, Akhila Akkala, 
Omnia Dafalla, and Amanda Wijesinghe.
