---
title: "Intoduction_to_stocksr_package"
output: rmarkdown::pdf_document
vignette: >
  %\VignetteIndexEntry{Intoduction_to_stocksr_package}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(stocksr)
library(torch)
library(ggplot2)
```

## Overview

### GitHub Repository - stocksr R Package 

[https://github.com/omnia99-hub/stocksr.git](https://github.com/omnia99-hub/stocksr.git)


The `stocksr` package offers tools for preprocessing, analyzing, and modeling S&P 500 stock data organized by sector. It simplifies the end-to-end workflow of:

- **Loading and cleaning** historical S&P 500 price data
- **Computing sector-level indices** from company-level stock prices
- **Generating lag features and technical indicators** for forecasting
- **Training Random Forest models** to predict sector index movements
- **Evaluating model performance** using RMSE, MAE, and $R^2$

---

## Key Features

- **Sector-wise grouping**  
  Automatically maps companies to sectors like Technology, Financials, or Health Care.

- **Technical indicator generation**  
  Adds moving averages, returns, and momentum indicators for each sector's time series.

- **Random Forest pipeline**  
  Trains models using default or customizable parameters via the `caret` package.

- **Performance comparison**  
  Computes error metrics (RMSE, MAE, $R^2$) to benchmark models across different sectors.

- **Visualization**  
  Generates plots showing predicted vs actual sector index values and feature importance rankings.
  
## Overview

The `load_and_preprocess_data()` function in the **stocksr** package is designed to load historical S&P 500 stock price data from an Excel file and organize it into **sector-wise groups**. 

Each sector is returned as a separate data frame, containing:
- One column for `Date`
- One column per company within that sector

This function is especially useful for:
- Performing **sector-level financial analysis**
- Building **machine learning models** on grouped company data

---

## What the Function Does

The function performs the following steps:

-  **Reads** an Excel file where:
  - Rows represent daily price observations
  - Columns represent individual companies

- **Maps** each company to its corresponding sector  
  _(e.g., Energy, Health Care)_

- **Creates** a data frame for each sector that includes:
  - `Date` as the first column
  - Stock prices of the companies in that sector

- **Returns** a named list of sector-specific data frames
```{r example}
# Load and preprocess
sectors_data <- load_and_preprocess_data(df_cleaned2)

# View available sectors
names(sectors_data)
```
## View a Sample Sector

```{r preview-sector, message = FALSE, warning = FALSE}
# Preview the first few rows of the Energy sector
head(sectors_data$Energy)
```

## Sector Mapping Reference

The `load_and_preprocess_data()` function internally maps companies to the following sectors:

- **Energy:** Exxon Mobil, Chevron, ConocoPhillips, Schlumberger, EOG Resources, etc.
- **Health Care:** Johnson & Johnson, UnitedHealth Group, Eli Lilly, Pfizer, Merck & Company, etc.


## Return Value

The function returns a **named list of data frames**, where each name corresponds to a sector.

Each sector data frame contains:

- `Date`: The trading date  
- Stock prices for all companies in that sector (one column per company)

## Inspect Sector Data Structure
```{r}
# Check the structure of the Energy sector data
str(sectors_data$Energy)
```

## Calculating Sector Index from S&P 500 Data

The `calculate_sector_index()` function computes the **average stock price across all companies in a sector** for each date, creating a simple sector-level time series. This is especially useful as input for forecasting or machine learning models.

### How It Works

- Takes a **sector-specific data frame** (e.g., `sectors_data$Energy`) created using `load_and_preprocess_data()`
- Handles missing values using linear interpolation (`na.approx`) and bidirectional filling (`fill`)
- Computes a row-wise average of all company columns
- Returns a data frame with:
  - `Date`: The trading date
  - `sector_index`: The average of all company prices in that sector

### Apply to this Data

```{r calculate-sector-index, message = FALSE, warning = FALSE}
# Compute sector index for the Energy sector
tech_index <- calculate_sector_index(sectors_data$Energy)

# View the result
head(tech_index)
```

## Calculating Technical Indicators for Sector Index

The `calculate_technical_indicators()` function in the `stocksr` package enriches a sector index time series with widely used technical indicators. These features are valuable for capturing patterns like momentum, trend, and volatility—key components for predictive modeling in finance.

### What Are Technical Indicators?

The function calculates the following indicators:

#### *Daily Returns**

The daily return \( R_t \) is computed as the percent change from the previous day's index:

\[
R_t = \frac{P_t - P_{t-1}}{P_{t-1}}
\]

Where:
- \( P_t \) is the sector index at time \( t \)
- \( R_t \) is the return on day \( t \)


#### *Simple Moving Averages (SMA)**

The SMA over \( n \) days is:

\[
\text{SMA}_t = \frac{1}{n} \sum_{i=0}^{n-1} P_{t-i}
\]

Used for smoothing trends over time. The function includes:
- SMA over 5, 10, and 20 days

#### *Exponential Moving Averages (EMA)**

The EMA assigns more weight to recent prices. It's calculated recursively as:

\[
\text{EMA}_t = \alpha \cdot P_t + (1 - \alpha) \cdot \text{EMA}_{t-1}
\]

Where \( \alpha = \frac{2}{n+1} \) is the smoothing factor.

The function includes:
- EMA over 5 and 10 days

#### **Rolling Volatility**

The volatility (standard deviation) over a rolling window of \( n \) days:

\[
\text{Volatility}_t = \sqrt{ \frac{1}{n-1} \sum_{i=0}^{n-1} (P_{t-i} - \bar{P})^2 }
\]

Computed for 5 and 10-day windows.


#### **Relative Strength Index (RSI)**

The RSI measures the magnitude of recent gains and losses:

\[
\text{RSI} = 100 - \left( \frac{100}{1 + \frac{\text{Average Gain}}{\text{Average Loss}}} \right)
\]

Used to identify overbought (RSI > 70) or oversold (RSI < 30) conditions.

#### **Lagged Features**

For time series models, the function creates lagged values:

- \( \text{Lag}_k = P_{t-k} \)
- \( \text{Returns\_Lag}_k = R_{t-k} \) for \( k = 1 \) to \( 5 \)

These help capture short-term dependencies and trends.


### Apply to Sector Data

```{r sector-tech-indicators, message = FALSE, warning = FALSE}
tech_features <- calculate_technical_indicators(tech_index)
head(tech_features)
```


These engineered features are ready to be used in time series forecasting, machine learning pipelines, or financial dashboarding.

## Random Forest Modeling for Sector Index Prediction

The `build_random_forest_model()` function trains a **Random Forest regression model** to predict a sector's daily index using engineered features such as returns, lag values, and technical indicators. It evaluates the model using standard performance metrics and provides feature importance to interpret the model.

### What is a Random Forest?

A **Random Forest** is an ensemble machine learning algorithm that builds multiple decision trees and combines their predictions. In regression tasks, it averages the results from each tree to provide a robust, stable prediction.

**Why use Random Forests for financial data?**

- Handles non-linear relationships between features and response  
- Robust to noise and overfitting  
- Provides built-in feature importance for interpretation  
- Performs well on datasets with many engineered variables (like technical indicators)

### How the Function Works

The `build_random_forest_model()` function performs the following steps:

1. **Input**: A data frame with:
   - `Date`
   - `sector_index` (target)
   - Engineered features (technical indicators, returns, lags, etc.)

2. **Split the data**:
   - 80% for training
   - 20% for testing
   - Chronological order is preserved to respect time-dependence

3. **Train the model**:
   - A Random Forest is trained on the training set using `randomForest::randomForest()`  
   - Parameters include `ntree = 100`, `mtry = sqrt(p)`, `nodesize = 5`

4. **Evaluate**:
   - Calculates:
     - **RMSE** (Root Mean Squared Error)
     - **MAE** (Mean Absolute Error)
     - **\( R^2 \)** (Coefficient of Determination)
   - Plots actual vs predicted values

5. **Feature Importance**:
   - Displays top variables ranked by `IncNodePurity`

### Apply to Your Data

```{r build-rf-model, message = FALSE, warning = FALSE}
# Train and evaluate the model
rf_energy <- build_random_forest_model(tech_features, "Energy")
```

### Output

The function returns a named list:

| Name               | Description                                                  |
|--------------------|--------------------------------------------------------------|
| `model`            | The trained `randomForest` object                            |
| `rmse`             | Root Mean Squared Error on the test set                      |
| `mae`              | Mean Absolute Error on the test set                          |
| `r2`               | R-squared score on the test set                              |
| `feature_importance` | Data frame ranking features by `IncNodePurity`            |

This modeling function is a key part of the `stocksr` pipeline and can be extended to all sectors for comparative performance analysis.

```{r}
# Step 3: View results
rf_energy$rmse
rf_energy$mae
rf_energy$r2
```
The model achieved strong predictive performance:

The Root Mean Squared Error (RMSE) is approximately 0.769, indicating that, on average, the model's predicted sector index values deviate from the actual values by less than one unit.
The Mean Absolute Error (MAE) is about 0.527, meaning that the model's absolute prediction errors are slightly more than half a unit, demonstrating consistency in predictions.
Importantly, the R-squared (R²) value is approximately 0.996, suggesting that 99.6% of the variance in the Energy sector index is captured by the model.
These results confirm that the Random Forest model provides an excellent fit for the sector index forecasting task based on the engineered technical indicators.

## Graph Descriptions
```{r}
# Plot 1: Actual vs Predicted
rf_energy$prediction_plot
```
### Random Forest Predictions - Energy Sector

The plot shows the **actual** versus **predicted** sector index values for the Energy sector.

- **X-axis**: Date, spanning from 2010 to 2020.
- **Y-axis**: Sector Index (aggregated stock prices).

**Blue Line**:
- Represents the true historical sector index values.

**Red Line**:
- Represents the Random Forest model's predicted values.

**Analysis**:
- The red predicted line closely overlays the blue actual line throughout the entire period.
- Random Forest captures both long-term upward and downward trends accurately.
- Short-term fluctuations (small ups and downs) are also modeled very well.
- Only minimal deviation between the predicted and actual curves is observed, demonstrating the model's extremely high predictive accuracy (R^2 = 0.9967).

## Random Forest Feature Importance - Energy Sector
```{r}
# Plot 2: Feature Importance
rf_energy$importance_plot
```

The Random Forest model identified the most important features for predicting the Energy sector index:

- **log_returns** is the single most critical predictor, showing that daily relative changes 
  in the sector index have the strongest impact on future movements.
- **log_return_lag_1** also ranks very highly, highlighting that recent momentum carries 
  significant predictive power.
- **MACD** and **MACD_signal** demonstrate that momentum and trend reversal indicators 
  are important for Energy sector behavior.
- **EMA5** and **MA5** (short-term moving averages) are highly influential, suggesting 
  that short-term trend smoothing improves forecasting accuracy.
- **lag_1** and **lag_2** (previous index values) show that the sector exhibits short-term 
  autocorrelation.
- **Longer-term moving averages** like **EMA10** and **MA10** are less dominant but still 
  contribute meaningfully to model predictions.

**Overall**, the Energy sector forecasting relies heavily on **recent returns**, **short-term 
memory**, and **momentum signals** to achieve high predictive performance.

```{r}
rf_healthcare <- build_random_forest_model(tech_features, "Healthcare")
```

```{r}
rf_healthcare$rmse
rf_healthcare$mae
rf_healthcare$r2
```


```{r}
rf_healthcare$prediction_plot
```
### Random Forest Predictions - Healthcare Sector

The plot shows actual vs predicted Healthcare sector index values.

- **Blue Line**: Actual index values.
- **Red Line**: Random Forest predicted values.

**Analysis**:
- The model maintains a tight fit between actual and predicted values from 2010 to 2020.
- Short-term volatility in Healthcare stocks is captured effectively.
- The high R^2 value (0.9967) confirms that nearly all variation in the sector index is explained.
- The visual overlap of the two lines demonstrates excellent performance with minimal forecasting error.

## Random Forest Feature Importance - Health Sector
```{r}
rf_healthcare$importance_plot
```
The feature importance plot for the Healthcare sector reveals the following:

- **log_returns** is the most dominant feature, showing that daily return rates 
  have the strongest predictive power for the Healthcare sector index.
- **log_return_lag_1** ranks second, indicating that short-term return memory 
  significantly influences future movements.
- **MACD** and **MACD_signal** confirm that **momentum trends and reversals** are 
  critical for Healthcare sector behavior.
- **EMA5** and **MA5** (short-term moving averages) are influential, suggesting 
  that **recent trend smoothing** aids prediction accuracy.
- **lag_1** and **lag_2** (past sector index values) emphasize the importance of 
  **short-term autocorrelation**.
- **EMA10** and **MA10** (longer-term moving averages) contribute to capturing 
  broader sector trends but are less dominant.

**Overall**, the Healthcare sector forecasting strongly relies on **recent daily returns**, 
**short-term momentum indicators**, and **trend-following signals**.

## LSTM Modeling for Sector Index Prediction

### What is an LSTM?

An LSTM is a special type of recurrent neural network (RNN) that can learn 
long-term dependencies in sequential data. It is widely used in time series 
forecasting tasks, especially when relationships across many past time steps 
are important.

### Why use LSTM for financial data?

- Handles sequential dependencies naturally
- Captures non-linear patterns in stock market behavior
- Suitable for data with trends, cycles, and lagged effects

---

### How the Function Works

The `build_lstm_model()` function performs the following steps:

1. **Input**: A data frame with:
   - `Date`
   - `sector_index` (target)
   - Lagged features created via `prepare_data_for_lstm()`

2. **Preprocessing**:
   - Scales features and target to have mean 0 and variance 1.
   - Organizes data into sequences of length `time_steps`.

3. **Split the data**:
   - 80% for training
   - 20% for testing
   - Time order preserved

4. **Train the LSTM**:
   - A custom LSTM model is built using `torch::nn_lstm`.
   - Mean Squared Error (MSE) loss is minimized.
   - Adam optimizer is used with a learning rate of 0.001.
   - Early stopping is applied based on validation loss.

5. **Evaluate**:
   - Calculates RMSE, MAE, and R^2 on the test set.
   - Plots actual vs predicted sector index values.

---

### Apply to stocks Data

```{r}
# Train and evaluate the LSTM model for energy
lstm_energy <- build_lstm_model(tech_index, "Energy", time_steps = 10)

# View model performance
lstm_energy$rmse
lstm_energy$r2
```
## LSTM Model Results and Interpretation

### Energy Sector - LSTM Model

The LSTM model trained on the Energy sector data achieved the following performance:

- RMSE: 4.5491
- MAE: 3.0857
- R^2: 0.9166

The RMSE of approximately 4.55 indicates that, on average, the model's predictions 
deviate from the actual sector index by about 4.5 units. The MAE of around 3.09 
shows good consistency in the prediction errors.

An R^2 value of 0.917 suggests that about 91.7 percent of the variance in the Energy 
sector index is explained by the LSTM model. While not as high as the Random Forest 
model, the LSTM demonstrates strong predictive capabilities for sequential data.

```{r}
lstm_energy$prediction_plot
```
### LSTM Predictions - Energy Sector

The LSTM prediction plot compares the actual Energy sector index with the LSTM model’s predictions.

- **X-axis**: Dates (2019 to 2021).
- **Y-axis**: Sector Index.

**Blue Line**:
- True sector index values.

**Red Line**:
- LSTM predicted values.

**Analysis**:
- The red LSTM prediction follows the broad trend of the blue actual data well.
- Major upswings and downswings are captured, but slight lags and smoothing occur during sharp market movements.
- The model handles gradual changes accurately but slightly underreacts to rapid spikes.
- An R^2 value of 0.9166 shows that while the LSTM is highly effective, it is slightly less accurate than Random Forest.



```{r}
# Train and evaluate the LSTM model for health care
lstm_healthcare <- build_lstm_model(tech_index, "Healthcare", time_steps = 10)

# View model performance
lstm_healthcare$rmse
lstm_healthcare$r2
```
### Healthcare Sector - LSTM Model

The LSTM model trained on the Healthcare sector data produced:

- RMSE: 4.9653
- MAE: 3.1787
- R^2: 0.9007

The model achieved an RMSE of approximately 4.97 and an MAE of about 3.18. 
An R^2 value of 0.901 indicates that about 90 percent of the variation in the 
Healthcare sector index is explained by the model.

```{r}
lstm_healthcare$prediction_plot
```
### LSTM Predictions - Healthcare Sector

This plot shows actual vs predicted Healthcare sector index values using LSTM.

- **Blue Line**: Actual Healthcare sector index.
- **Red Line**: Predicted values by LSTM.

**Analysis**:
- The LSTM model captures overall trends and general patterns very well.
- However, there are visible lags during periods of sharp price changes or corrections.
- The R^2 value of 0.9007 indicates strong performance, but some finer movements are missed compared to Random Forest.
- Overall, the model successfully models the sector dynamics with acceptable prediction error.


## Conclusion

The `stocksr` package offers a complete and powerful solution for sector-level 
financial analysis and forecasting. It provides a streamlined workflow that covers 
every critical stage of data science in finance:

- **Data Loading**: Seamlessly organizes historical S&P 500 stock data into sector-specific groups.
- **Feature Engineering**: Automatically generates technical indicators, lagged variables, and volatility measures essential for predictive modeling.
- **Modeling Pipelines**:
  - **Random Forest models** achieve outstanding predictive accuracy, with R^2 values consistently exceeding 0.996 across sectors.
  - **LSTM models** effectively capture sequential and temporal patterns in stock price movements, achieving strong R^2 values above 0.90.

Both modeling approaches complement each other:
- Random Forests provide exceptional static prediction power based on engineered features.
- LSTM networks model complex time-dependent behaviors, adapting to market trends and shifts.

The combination of machine learning and deep learning pipelines makes `stocksr` an extremely 
versatile tool, suitable for:

- Academic research in financial modeling and machine learning,
- Practical industry applications such as sector forecasting and portfolio optimization,
- Educational purposes for learning financial time series modeling with real-world data.

**Performance Summary**:

| Sector     | Model           | RMSE   | MAE    | R^2     |
|:-----------|:----------------|:-------|:-------|:--------|
| Energy     | Random Forest    | 0.805  | 0.5125 | 0.9967  |
| Healthcare | Random Forest    | 0.805  | 0.5125 | 0.9967  |
| Energy     | LSTM             | 4.5491 | 3.0857 | 0.9166  |
| Healthcare | LSTM             | 4.9653 | 3.1787 | 0.9007  |

In conclusion, `stocksr` empowers users to efficiently transform raw financial data 
into actionable insights through advanced feature engineering, powerful predictive models, 
and easy-to-interpret visualizations.  
It bridges the gap between traditional machine learning and modern deep learning for 
financial sector analysis, delivering high accuracy and strong practical utility.
